---
title: "Introduction to RegionalClassification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to RegionalClassification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

The `RegionalClassification` package provides tools to convert Belgian geographic codes between classification systems and across time.

Belgian administrative data can reference geography in many ways — NIS commune codes, NUTS regions, postal codes, internal arrondissement codes — and these change over time as communes merge or district borders shift. This package unifies all of them.

## Loading the package

```{r load}
source("main.R")
# master_data is now available — it contains all lookup tables
```

On load, the package reads all source files from `data/raw/` and builds a unified master table. You'll see messages like:

```
Master table built: 583 communes NIS 2019, 567 communes NIS 2025
  Postal codes NIS 2019: 1149, NIS 2025: 1148
  NIS changes 2019->2025: 30 entries
  NIS BEFORE_2019: 589 communes
```

---

## 1. Basic conversions

The main function is `convert_codes()`:

```{r}
convert_codes(codes, from, to, master_data)
```

### NIS communes → NUTS3

```{r}
result <- convert_codes(
  c(21004L, 11002L, 44021L, 62063L),
  from = "NIS_COMMUNE_2019",
  to   = "NUTS3_2021",
  master_data
)
# code_from  code_to
#     21004   BE100     (Bruxelles-Capitale)
#     11002   BE211     (Antwerpen)
#     44021   BE234     (Arr. Gent)
#     62063   BE332     (Arr. Liège)
```

### Postal codes → NIS communes

```{r}
convert_codes(c(1000L, 2000L, 4000L), "POSTAL", "NIS_COMMUNE_2019", master_data)
```

### NIS commune → all geographic levels

```{r}
commune <- 11002L  # Antwerpen

convert_codes(commune, "NIS_COMMUNE_2019", "NIS_ARRONDISSEMENT_2019", master_data)
convert_codes(commune, "NIS_COMMUNE_2019", "NIS_PROVINCE_2019",       master_data)
convert_codes(commune, "NIS_COMMUNE_2019", "NIS_REGION_2019",         master_data)
convert_codes(commune, "NIS_COMMUNE_2019", "NUTS3_2021",              master_data)
convert_codes(commune, "NIS_COMMUNE_2019", "NUTS2_2021",              master_data)
convert_codes(commune, "NIS_COMMUNE_2019", "INTERNAL_ARRONDISSEMENT", master_data)
```

---

## 2. NIS versions

Belgian communes are periodically reorganized. Three versions are supported:

```
BEFORE_2019  ──►  2019  ──►  2025
  591 comm.       583         567
```

### Converting across NIS versions

```{r}
# NIS 2019 → NIS 2025 (handles fusions)
convert_codes(c(71002L, 71011L), "NIS_COMMUNE_2019", "NIS_COMMUNE_2025", master_data)

# NIS 2025 → NIS 2019 (fused commune returns multiple rows)
convert_codes(71072L, "NIS_COMMUNE_2025", "NIS_COMMUNE_2019", master_data)

# NIS BEFORE_2019 → NIS 2019
convert_codes(c(55022L, 56011L), "NIS_COMMUNE_BEFORE_2019", "NIS_COMMUNE_2019", master_data)
```

### Historical NUTS assignments

BEFORE_2019 communes are assigned their NUTS codes as of 31 December 2018:

```{r}
# Get NUTS3 (2021) for a pre-2019 commune
convert_codes(55022L, "NIS_COMMUNE_BEFORE_2019", "NUTS3_2021", master_data)

# Get NUTS3 (2027) for a pre-2019 commune
convert_codes(55022L, "NIS_COMMUNE_BEFORE_2019", "NUTS3_2027", master_data)
```

---

## 3. NUTS 2027

NUTS 2027 codes are derived from NUTS 2021 per EU regulation 2026/195. The main changes for Belgium affect Antwerpen and the Flemish provinces.

```{r}
# NIS 2019 communes → NUTS 2027
convert_codes(
  c(21004L, 11002L, 44021L),
  "NIS_COMMUNE_2019", "NUTS3_2027",
  master_data
)
# 21004 → BE100  (Brussels, unchanged)
# 11002 → BE261  (Antwerpen: BE211 → BE261)
# 44021 → BE274  (Gent: BE234 → BE274)
```

### NUTS 2021 ↔ NUTS 2027 roundtrip

```{r}
nuts_2021 <- c("BE211", "BE223", "BE231", "BE335")

result_2027 <- convert_codes(nuts_2021, "NUTS3_2021", "NUTS3_2027", master_data)
# BE211 → BE261, BE223 → BE226, BE231 → BE271, BE335 → BE335 (unchanged)

result_back <- convert_codes(result_2027$code_to, "NUTS3_2027", "NUTS3_2021", master_data)
# BE261 → BE211, BE226 → BE223, BE271 → BE231, BE335 → BE335
```

---

## 4. Ambiguous conversions

Some conversions are inherently M:N. The arrondissement of Verviers (63000) spans two NUTS3 regions: BE335 (francophone) and BE336 (germanophone).

```{r}
# Will raise an error by default
convert_codes(63000L, "NIS_ARRONDISSEMENT_2019", "NUTS3_2021", master_data)
# Error: Conversion from '...' to '...' is NOT a simple (direct) conversion.

# Force all possible mappings
result <- convert_codes(63000L, "NIS_ARRONDISSEMENT_2019", "NUTS3_2021",
                        master_data, allow_ambiguous = TRUE)
# code_from  code_to
#     63000   BE335
#     63000   BE336
```

### Checking a path before converting

```{r}
check_conversion_path("NIS_ARRONDISSEMENT_2019", "NUTS3_2021")
# $is_simple    FALSE
# $explanation  "... ambiguous steps (Verviers) ..."

check_conversion_path("NIS_COMMUNE_2019", "NUTS3_2021")
# $is_simple    TRUE
# $path         "NIS_COMMUNE_2019 -> NUTS_LAU_2021 -> NUTS3_2021"
```

---

## 5. Fuzzy name matching

Match approximate or misspelled names to official codes:

```{r}
fuzzy_match_names(
  names    = c("Bruxeles", "Antwerpn", "Liege", "Vervirs"),
  target   = "NIS_COMMUNE_2019",
  master_data,
  max_dist = 0.3,
  language = "both"   # "fr", "nl", or "both"
)
#   input_name  matched_name  matched_code  distance  is_confident
#   Bruxeles    Bruxelles        21004       0.037      TRUE
#   Antwerpn    Antwerpen        11002       0.037      TRUE
#   Liege       Liège            62063       0.150      TRUE
#   Vervirs     Verviers         63079       0.042      TRUE
```

---

## 6. Available conversions

```{r}
# List all supported conversion routes
list_available_conversions()

# Print full path info for a specific conversion
print_conversion_check("POSTAL", "NUTS3_2021")
```

---

## 7. Classification identifiers

The `from` and `to` arguments accept several aliases:

| Standard | Aliases |
|---|---|
| `NIS_COMMUNE_2019` | `NIS_COM_2019`, `COMMUNE_2019` |
| `NIS_COMMUNE_2025` | `NIS_COM_2025`, `COMMUNE_2025` |
| `NIS_COMMUNE_BEFORE_2019` | `NIS_COM_BEFORE_2019` |
| `NUTS3_2021` | `NUTS_2021` |
| `NUTS3_2027` | `NUTS_2027` |
| `INTERNAL_ARRONDISSEMENT` | `INTERNAL`, `INTERNAL_ARR`, `INTERNE` |
| `POSTAL` | `CODE_POSTAL`, `CP` |

---

## 8. Pending features

The following features are ready in the code but require additional data files not yet available:

| Feature | Waiting for |
|---|---|
| `NIS_COMMUNE_2025` → `NUTS3_2027` | `CONVERSION_NIS2025_NUTS2027.xlsx` |

Once the file is placed in `data/raw/` and the package reloaded, the conversion will work automatically.
