---
title: "Getting started with nbbbenuts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with nbbbenuts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

The `nbbbenuts` package converts Belgian geographic codes between classification
systems and across time.

Belgian administrative data references geography in many ways — NIS commune codes,
NUTS regions, postal codes, internal arrondissement codes — and these change over
time as communes merge or district borders shift. This package unifies all of them.

All classification identifiers are strict uppercase strings (e.g. `"NIS_MUNICIPALITY_2019"`).
See `?classification_reference` for the full list, or call:

```{r}
get_all_classification_nodes()
```

---

## Loading the package

```{r load}
library(nbbbenuts)
master_data <- load_master_data()
# Master data loaded from '.../inst/extdata':
# 583 communes NIS 2019, 567 NIS 2025, 589 NIS BEFORE_2019
```

To rebuild the snapshot after updating raw source files in `data/raw/`:

```{r rebuild, eval=FALSE}
master_data <- rebuild_master_data()   # requires data/raw/ files + readxl
```

---

## 1. Converting codes

The main function is `convert_codes()`:

```{r}
convert_codes(codes, from, to, master_data)
```

### NIS communes → NUTS3

```{r}
result <- convert_codes(
  c(21004L, 11002L, 44021L, 62063L),
  from = "NIS_MUNICIPALITY_2019",
  to   = "NUTS_DISTRICT_2021",
  master_data
)
# code_from  code_to
#     21004   BE100     (Bruxelles-Capitale)
#     11002   BE211     (Antwerpen)
#     44021   BE234     (Arr. Gent)
#     62063   BE332     (Arr. Liège)
```

### Postal codes → NIS communes

```{r}
convert_codes(c(1000L, 2000L, 4000L), "POSTAL", "NIS_MUNICIPALITY_2019", master_data)
```

### NIS commune → all geographic levels

```{r}
commune <- 11002L  # Antwerpen

convert_codes(commune, "NIS_MUNICIPALITY_2019", "NIS_DISTRICT_2019", master_data)
convert_codes(commune, "NIS_MUNICIPALITY_2019", "NIS_PROVINCE_2019",       master_data)
convert_codes(commune, "NIS_MUNICIPALITY_2019", "NIS_REGION_2019",         master_data)
convert_codes(commune, "NIS_MUNICIPALITY_2019", "NUTS_DISTRICT_2021",              master_data)
convert_codes(commune, "NIS_MUNICIPALITY_2019", "NUTS_PROVINCE_2021",              master_data)
convert_codes(commune, "NIS_MUNICIPALITY_2019", "NBB_DISTRICT_2021", master_data)
```

---

## 2. NIS versions

Belgian communes are periodically reorganized. Three versions are supported:

```
BEFORE_2019  -->  2019  -->  2025
  591 comm.       583         567
```

Temporal conversions return a `nature` column describing the type of change:

| `nature`       | Meaning |
|----------------|---------|
| `UNCHANGED`    | Code persists unchanged (553 communes in 2019->2025) |
| `FUSION`       | One or more communes merged into a single code |
| `CHANGE_DSTR`  | Commune moved to a different arrondissement (2 communes) |
| `CHANGE_PROV`  | Commune moved to a different province (1 commune) |

```{r}
# NIS 2019 -> NIS 2025: nature column shows what changed
convert_codes(c(71002L, 71011L), "NIS_MUNICIPALITY_2019", "NIS_MUNICIPALITY_2025", master_data)
#    code_from  code_to   nature
#        71002    71072   FUSION
#        71011    71072   FUSION

# CHANGE_DSTR: commune 44045 moved to a different arrondissement
convert_codes(44045L, "NIS_MUNICIPALITY_2019", "NIS_MUNICIPALITY_2025", master_data)
#    code_from  code_to       nature
#        44045    46029  CHANGE_DSTR

# NIS 2025 -> NIS 2019 (fused commune returns multiple rows)
convert_codes(71072L, "NIS_MUNICIPALITY_2025", "NIS_MUNICIPALITY_2019", master_data)

# NIS BEFORE_2019 -> NIS 2019
convert_codes(c(55022L, 56011L), "NIS_MUNICIPALITY_BEFORE_2019", "NIS_MUNICIPALITY_2019", master_data)
```

---

## 3. NUTS 2027

NUTS 2027 codes for NIS 2025 communes come from the official Statbel file. For
NIS 2019 and BEFORE_2019 communes, NUTS 2027 is derived by remapping NUTS 2021
codes per EU regulation 2026/195.

```{r}
convert_codes(
  c(21004L, 11002L, 44021L),
  "NIS_MUNICIPALITY_2019", "NUTS_DISTRICT_2027",
  master_data
)
# 21004 → BE100  (Brussels, unchanged)
# 11002 → BE261  (Antwerpen: BE211 → BE261)
# 44021 → BE274  (Gent: BE234 → BE274)
```

---

## 4. Ambiguous conversions

Some conversions are inherently M:N. The arrondissement of Verviers (63000) spans
two NUTS3 regions: BE335 (francophone) and BE336 (germanophone).

```{r}
# Force all possible mappings
convert_codes(63000L, "NIS_DISTRICT_2019", "NUTS_DISTRICT_2021",
              master_data, allow_ambiguous = TRUE)
# code_from  code_to
#     63000   BE335
#     63000   BE336

# Check a path before converting
check_conversion_path("NIS_DISTRICT_2019", "NUTS_DISTRICT_2021")
# $is_simple    FALSE
```

---

## 5. Dataset conversion

Convert a whole column in a `data.frame` or `data.table`:

```{r}
library(data.table)
dt <- data.table(
  commune = c(21004L, 11002L, 44021L),
  value   = c(100, 200, 300)
)

convert_dataset(dt, "commune",
                from = "NIS_MUNICIPALITY_2019",
                to   = "NUTS_DISTRICT_2021",
                master_data)
# Adds column 'cd_nuts3_2021' with converted codes.
```

---

## 6. Validating and labelling codes

### Validate codes

Check whether codes belong to a known classification:

```{r}
validate_codes(c(21004L, 99999L, 11002L), "NIS_MUNICIPALITY_2019", master_data)
#    code  is_valid
#   21004      TRUE
#   99999     FALSE
#   11002      TRUE
```

### Get official names

```{r}
get_label(c(21004L, 11002L), "NIS_MUNICIPALITY_2019", master_data, lang = "fr")
#    code       label
#   21004   Bruxelles
#   11002      Anvers

get_label(c("BE100", "BE211"), "NUTS_DISTRICT_2021", master_data, lang = "nl")
```

### Build a crosswalk table

```{r}
# Simple pair: one row per source code
get_crosswalk("NIS_MUNICIPALITY_2019", "NUTS_DISTRICT_2021", master_data)

# With weights for the ambiguous Verviers split
get_crosswalk("NIS_DISTRICT_2019", "NUTS_DISTRICT_2021", master_data,
              weights = TRUE)
# NIS_DISTRICT_2019  NUTS_DISTRICT_2021  weight
#                   63000       BE335   0.500
#                   63000       BE336   0.500
# (equal weights by default; use register_split_weights() for custom ones)
```

---

## 7. Rebasing longitudinal data

`rebase_series()` converts a panel dataset that spans multiple classification
versions onto a single target version.

### N:1 merge (communes fused over time)

Communes 11002 and 11007 were merged into 11002 in NIS 2025:

```{r}
panel <- data.table(
  year      = c(2022L, 2022L, 2022L, 2025L, 2025L),
  commune   = c(11002L, 11007L, 21004L, 11002L, 21004L),
  population = c(18000, 8500, 180000, 28000, 185000)
)

rebase_series(
  panel,
  period_col  = "year",
  code_col    = "commune",
  value_cols  = "population",
  version_map = list("NIS_MUNICIPALITY_2019" = 2022L,
                     "NIS_MUNICIPALITY_2025" = 2025L),
  to          = "NIS_MUNICIPALITY_2025",
  master_data = master_data
)
# 2022: commune 11002 → 18000 + 8500 = 26500 (fused)
# 2025: unchanged (already in NIS 2025)
```

### 1:N split with custom weights

For ratio variables (`value_type = "ratio"`), values are replicated unchanged to
each target region. For additive variables (the default), values are distributed
by weight.

```{r}
# Step 1: get the template with equal weights as a starting point
tpl <- split_weights_template("NIS_DISTRICT_2019", "NUTS_DISTRICT_2021", master_data)
#   code_from  code_to  weight
#       63000    BE335     0.5
#       63000    BE336     0.5

# Step 2: fill in population-based weights
tpl[code_from == "63000" & code_to == "BE335", weight := 0.857]
tpl[code_from == "63000" & code_to == "BE336", weight := 0.143]

# Step 3a: pass directly (one-off)
arr_data <- data.table(year = 2022L, arr = 63000L, emploi = 120000)

rebase_series(
  arr_data,
  period_col  = "year",
  code_col    = "arr",
  value_cols  = "emploi",
  version_map = list("NIS_DISTRICT_2019" = 2022L),
  to          = "NUTS_DISTRICT_2021",
  master_data = master_data,
  split       = tpl
)

# Step 3b: register for repeated use in the session
register_split_weights(
  "NIS_DISTRICT_2019", "NUTS_DISTRICT_2021", tpl, variable = "population"
)

rebase_series(
  arr_data,
  period_col  = "year",
  code_col    = "arr",
  value_cols  = "emploi",
  version_map = list("NIS_DISTRICT_2019" = 2022L),
  to          = "NUTS_DISTRICT_2021",
  master_data = master_data
  # split = "population" is the default
)
```

---

## 8. Diagnosing code coverage

`diagnose_classification()` verifies alignment between your data and a reference
classification, or auto-detects the best match.

### Check mode

```{r}
nuts3_data <- data.table(
  nuts3 = c("BE100", "BE211", "BE332"),
  value = c(100, 200, 150)
)

diagnose_classification(nuts3_data, "nuts3", master_data,
                        classification = "NUTS_DISTRICT_2021")
# ================================================================
#   CLASSIFICATION DIAGNOSTIC
# ================================================================
#   Dataset        : 3 rows  |  column 'nuts3'
#   Classification : NUTS3
#   Version        : 2021
#   Reference      : NUTS_DISTRICT_2021  (44 codes)
# ================================================================
#   Coverage   : 3 / 44  [##------------------] 6.8%
#   ...
#   Status: !!  INCOMPLETE -- missing reference codes
```

The return value is an invisible list with `coverage_rate`, `n_missing`,
`n_unknown`, `n_duplicates`, `missing_codes`, `unknown_codes`, and `status`.

### Auto-detect mode

```{r}
diagnose_classification(nuts3_data, "nuts3", master_data)
# Ranks all 22 classifications by match rate and recommends the best one.
```

---

## 9. Fuzzy name matching

Match approximate or misspelled names to official codes:

```{r}
fuzzy_match_names(
  names    = c("Bruxeles", "Antwerpn", "Liege", "Vervirs"),
  target   = "NIS_MUNICIPALITY_2019",
  master_data,
  max_dist = 0.3,
  language = "both"
)
#   input_name  matched_name  matched_code  distance  is_confident
#   Bruxeles    Bruxelles        21004       0.037      TRUE
#   Antwerpn    Antwerpen        11002       0.037      TRUE
#   Liege       Liège            62063       0.150      TRUE
#   Vervirs     Verviers         63079       0.042      TRUE
```

---

## 10. All classification identifiers

```{r}
# See the full list
get_all_classification_nodes()

# Detailed help page with conversion table
?classification_reference
```

Key identifiers:

| Level            | Identifiers |
|------------------|-------------|
| Communes         | `NIS_MUNICIPALITY_BEFORE_2019`, `NIS_MUNICIPALITY_2019`, `NIS_MUNICIPALITY_2025` |
| Arrondissements  | `NIS_DISTRICT_2019`, `NIS_DISTRICT_2025` |
| Provinces        | `NIS_PROVINCE_2019`, `NIS_PROVINCE_2025` |
| Regions          | `NIS_REGION_2019`, `NIS_REGION_2025` |
| NUTS 2021        | `NUTS_MUNICIPALITY_2021`, `NUTS_DISTRICT_2021`, `NUTS_PROVINCE_2021`, `NUTS_REGION_2021`, `NUTS_COUNTRY` |
| NUTS 2027        | `NUTS_DISTRICT_2027`, `NUTS_PROVINCE_2027`, `NUTS_REGION_2027` |
| Other            | `POSTAL`, `NBB_DISTRICT_2021` |
